---
title: "Data Exploration in R"
subtitle: "Heatmaps"
author: "Prof. Dr. Jan Kirenz"
institute: "HdM Stuttgart"
output:
  xaringan::moon_reader:
    css: ["footer-header.css", "default", "default-fonts"]
    lib_dir: libs
    nature:
      titleSlideClass: [center, middle]
      highlightLines: true
      highlightStyle: github
      countIncrementalSlides: false
---
layout: true
  
<div class="my-footer"><span> Data Exploration in R | Prof. Dr. Jan Kirenz </span></div> 

```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, dpi = 300)

library(tidyverse)
library(gt)
library(flair)
library(kableExtra)


```

---

## Heatmap

- In this presentation you will learn how to perform data exploration using heatmaps. 


<div class = "blue">

You can use this kind of heatmap if the variable is: 

<p>

- Categorical,  <br>
- discrete with only a few values or <br>
- continuous

</div>


---


## Data

- We use the dataset `wage` 
- It contains wage and other data for a group of 3000 male workers.

```{r}
library(tidyverse)

wage_df <- read_csv("https://raw.githubusercontent.com/kirenz/datasets/master/wage.csv")

```

- The data includes 3000 observations on 11 variables. 


---

## Two categorical variables

- Heatmaps for categorical and discrete Variables 
- We use the function [geom_bin2d()](https://ggplot2.tidyverse.org/reference/geom_bin2d.html) from the ggplot2 package.


```{r heat-plot, echo=FALSE, fig.cap="Simple Heatmap", out.width = "60%"}

wage_df %>%
  ggplot(aes(jobclass, education)) +
  geom_bin2d() +
  scale_fill_gradient(low = "gray85", high = "steelblue") +
  labs(fill = "Workers", y = "Education", x = "Job class") +
  theme_classic() 

```

---

## Two categorical variables

```{r heat-code, eval=FALSE}

wage_df %>%
  ggplot(aes(jobclass, education)) +
  geom_bin2d() + #<<
  scale_fill_gradient(low = "gray85", high = "steelblue") +
  labs(fill = "Workers", y = "Education", x = "Job class") +
  theme_classic() 

```

- `geom_bin2d()` divides the plane into rectangles, counts the number of cases in each rectangle, and then (by default) maps the number of cases to the rectangle's fill.

---

## Combination of Variables

- Combination of categorical and continous   variables


```{r, echo = FALSE, out.width = "60%"}

wage_df %>% 
  ggplot(aes(jobclass, education, fill = wage )) +
  geom_tile() +
  scale_fill_gradient(low = "gray85", high = "steelblue")  +
  theme_classic()

```

---


## Combination of Variables

- Combination of categorical and continous   variables


```{r, eval = FALSE}

wage_df %>% 
  ggplot(aes(jobclass, education, fill = wage )) +
  geom_tile() +
  scale_fill_gradient(low = "gray85", high = "steelblue")  +
  theme_classic()

```

---


class: center, middle

![](https://upload.wikimedia.org/wikipedia/commons/5/5e/HdM_Logo.svg)

## Thanks!

[Prof. Dr. Jan Kirenz](https://www.kirenz.com)  

HdM Stuttgart  
Nobelstraße 10  
70569 Stuttgart
