---
title: "Data Exploration in R"
subtitle: "Basics"
author: "Prof. Dr. Jan Kirenz"
institute: "HdM Stuttgart"
output:
  xaringan::moon_reader:
    css: ["footer-header.css", "default", "default-fonts"]
    lib_dir: libs
    nature:
      titleSlideClass: [center, middle]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
layout: true
  
<div class="my-footer"><span> Data Exploration in R | Prof. Dr. Jan Kirenz </span></div> 

```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, dpi = 300)

library(tidyverse)
library(gt)
library(flair)


```

---

## Libraries

To use the code in this presentation, activate the following packages: 

```{r}
library(tidyverse)
library(gt)
```

---

## Data

- We use the dataset `wage` 
- It contains wage and other data for a group of 3000 male workers.

```{r}
library(tidyverse)

wage_df <- read_csv("https://raw.githubusercontent.com/kirenz/datasets/master/wage.csv")

```

- The data includes 3000 observations on 11 variables. 


---

## Data


.pull-left[

* **year**: Year that wage information was recorded

* **age**: Age of worker

* **maritl**: A factor with levels: 

 1. Never Married 
 2. Married 
 3. Widowed 
 4. Divorced and 
 5. Separated indicating marital status

* **race**: A factor with levels: 

 1. White 
 2. Black 
 3. Asian and 
 4. Other indicating race

]

.pull-right[

* **education**: A factor with levels: 

 1. < HS Grad 
 2. HS Grad 
 3. Some College 
 4. College Grad and 
 5. Advanced Degree indicating education level

* **jobclass**: A factor with levels: 

 1. Industrial and 
 2. Information indicating type of job

* **logwage**: Log of workers wage

* **wage**: Workers raw wage

]
---

## Counts and Tables

<div class = "blue">

You should use this method if the data is: Categorical

</div>


In this section you will learn how to do data exploration for categorical variables using tables (also called contingeny tables) and counts. 

---

## Count


.pull-left[

* Use data `wage_df`.
* Perform [count()](https://dplyr.tidyverse.org/reference/count.html) on `maritl` 
* Sort the values. 
* Use [gt()](https://gt.rstudio.com) to print the table.

]

.pull-right[

```{r count-one, include = FALSE}

wage_df %>% 
  count(maritl,
        sort = TRUE) %>% 
  gt()

```


```{r count-one-color, echo=FALSE}

decorate("count-one") %>%
  mask("wage_df") %>%
  mask("count") %>%
  mask("maritl") %>%
  mask("sort") %>%
  mask("gt") %>%
  knit_print.with_flair()

```
]

---

## Count

.pull-left[

* Use data `wage_df`.
* Perform [count()](https://dplyr.tidyverse.org/reference/count.html) on `maritl` and `education`
* Sort the values. 
* Use [gt()](https://gt.rstudio.com) to print the table.

]

.pull-right[

Count the combined appearences of `maritl` and `education` and sort the values:

```{r count-two, include=FALSE}

wage_df %>% 
  count(maritl, education,
        sort= TRUE) %>% 
  gt()

```


```{r count-two-color, echo=FALSE}

decorate("count-two") %>%
  mask("wage_df") %>%
  mask("count") %>%
  mask("maritl") %>%
  mask("education") %>%
  mask("sort") %>%
  mask("gt") %>%
  knit_print.with_flair()

```

]
---

## Count

.pull-left[

* Use data `wage_df`.
* Obtain the sum of the quantitative variable `wage` for the different levels of `maritl`
* Name the new variable "Sum". 
* Use gt() to print the table.

Read the [documentation for count()](https://dplyr.tidyverse.org/reference/count.html) to learn how to perform this task.
]

.pull-right[

```{r count-numeric, include=FALSE}

wage_df %>%  
  count(maritl,
        wt = wage,
        name = "Sum") %>% 
  gt()

```

```{r count-numeric-color, echo=FALSE}

decorate("count-numeric") %>%
  mask("wage_df") %>%
  mask("count") %>%
  mask("maritl") %>%
  mask("wt") %>%
  mask("name") %>%
  mask("gt") %>%
  knit_print.with_flair()

```


]
---

## Total counts



* Total counts are an useful way to represent the observations that fall into each combination of the levels of categorical variables. 

* We create a contingency table of the two categorical variables `jobclass` and `race` and call the result `tab`:



```{r tab}

tab <- table(wage_df$jobclass, wage_df$race)

kable(tab)

```


---

## Joint proportions

* We can also view the percentage of each cell in relation to the total amount of all observations (here n = 3000). 

* Therefore, you have to simply divide the numbers from our total counts with 3.000.

The following code generates tables of *joint* proportions:

```{r prop-tab}

prop <- prop.table(tab)*100

kable(prop, digits = 2)

```


For example, around .red[??] % of all people in the dataset are white industrial workers.

???

44%

---

## Conditional proportions

- You also may want to know the probability that workers have a certain jobclass, given that they have a particular ethnical background. 

- This is a so called **conditional probability**. 

- Conditional probability represents the chance that one event will occur given that a second event has already occurred. 


---

## Conditional proportions


* The following code generates tables of *conditional* proportions:

```{r prop-tab-col}

# conditional on columns
prop_col <- prop.table(tab, 2)*100  

kable(prop_col, digits = 2)

```



We performed a columnwise evaluation and are now able to answer the following question: 

* Approximately what proportion of all white workers are industrial workers? The answer is: around .red[??] %.

???

53%

---

## Conditional proportions: rows

Now we want to obtain the probability that workers have a certain race, given their jobclass. 


```{r prop-tab-row}

# conditional on rows
prop.table(tab, 1)  

```

We performed a rowwise evaluation and are now able to answer the following question: 

* Approximately what proportion of all industrial workers are white? 
* The answer is: around 86%. 

---

## Chi-squared Test of Independence

Finally, let's test the hypothesis whether the variable `jobclass` is independent of the variable `race` at .05 significance level.


```{r chi-sq}

chisq.test(tab)  

```

As the p-value is smaller than the .05 significance level, we reject the null hypothesis that the jobclass is independent of the race of the workers. 

---

class: center, middle

![](https://upload.wikimedia.org/wikipedia/commons/5/5e/HdM_Logo.svg)

## Thanks!

[Prof. Dr. Jan Kirenz](https://www.kirenz.com)  

HdM Stuttgart  
Nobelstraße 10  
70569 Stuttgart
